Automatic feature classification in stochastic block models

### 2021-04-15 nips/submission.tex
More inference fleshing out

%custom commands
\newcommand{\Xcal}{\mathcal{X}}
\newcommand{\Integers}{\mathbb{Z}}
\newcommand{\one}{\mathbbm{1}}
\newcommand{\Gaussian}{\mathcal{N}}
\newcommand{\indep}{\perp \!\!\! \perp}
\newcommand{\specialchoose}{\genfrac{\{}{\}}{0pt}{}}

@@

We are presented with a vertex-labelled graph $(A, X)$. The goal is to draw samples from equation \ref{eqn:theta-target}. However, this is not easily done in practice.

\begin{equation}
	\label{eqn:theta-target}
	\theta^{(i)} \sim p(\theta | A, X)
\end{equation}
%
We instead propose an iterative approach. First drawing samples from the block membership posterior $p(b | A, X)$ and then using these samples to draw samples from $\theta$ conditioned on the block membership sample $b^{(i)}$ and feature matrix. 
%
\begin{align}
	b^{(i)} &\sim p \Big( b | A, X \Big) \\
	\theta^{(i)} &\sim p\Big(\theta | X, b^{(i)} \Big)
\end{align}
%
Both of these can be implemented using a variation of the Metropolis-Hastings algorithm \cite{hastings-alg}. We just need to define a proposal distribution $q(\cdot, \cdot)$ for proposing moves and be able to evaluate the target distribution point-wise.

\subsection{Inferring block memeberships from graph $p (b | A, X)$}

\citet{Peixoto-MCMC} proposes a Monte Carlo method which we will base our approach on. It relies on writing the posterior in the following form.

\begin{equation}
	p(b | A, X) \propto p(A | b, X) \cdot p(b | X) \coloneqq \pi_b(b)
\end{equation}

>>>>

%custom commands
\newcommand{\Xcal}{\mathcal{X}}
\newcommand{\Bcal}{\mathcal{B}}
\newcommand{\Integers}{\mathbb{Z}}
\newcommand{\one}{\mathbbm{1}}
\newcommand{\Gaussian}{\mathcal{N}}
\newcommand{\indep}{\perp \!\!\! \perp}
\newcommand{\specialchoose}{\genfrac{\{}{\}}{0pt}{}}
\newcommand{\Expect}{\mathbb{E}}

@@

We are presented with a vertex-labelled graph $(A, X)$. The goal is to draw samples from equation \ref{eqn:theta-target}. However, this is not easily done in practice.
%
\begin{equation}
	\label{eqn:theta-target}
	\theta^{(i)} \sim p(\theta | A, X)
\end{equation}
%
We instead propose an iterative approach. First drawing samples $b^{(i)}$ from the block membership posterior (equation \ref{eqn:b-samples}). We then use each $b^{(i)}$ to draw samples for $\theta$ as in equation \ref{eqn:theta-samples}. 
%
\begin{align}
	b^{(i)} &\sim p \Big( b | A, X \Big)  \label{eqn:b-samples}\\
	\theta^{(i)} &\sim p\Big(\theta | X, b^{(i)} \Big) \label{eqn:theta-samples}
\end{align}
%
Both of these can be implemented with a Markov Chain through the Metropolis-Hastings algorithm \cite{hastings-alg}. We just need to define a proposal distribution $q(x, y)$ for proposing a move $x \rightarrow y$ and be able to evaluate an un-normalised form of the target distribution, denoted $\pi(\cdot)$, point-wise. The proposed move is then accepted with acceptance probability $\alpha$ else it is rejected and we stay at $x$.
%
\begin{equation}
	\alpha = \min \left( \frac{\pi(y) q(y, x)}{\pi(x) q(x, y)} , 1 \right)
\end{equation}
%
This accept-reject step ensures the resulting Markov Chain is in detailed balance with the target distribution $\pi(\cdot)$. What we propose in equations \ref{eqn:b-samples} and \ref{eqn:theta-samples} is therefore implemented through a 2-level Markov chain. The resulting samples for $\theta^{(i)}$ are unbiased in the sense that the expectation of their distribution is the posterior we are targeting in equation \ref{eqn:theta-target}.
%
\begin{align*}
	\Expect_{b^{(i)}} \left[p \left( \theta | X, b^{(i)} \right) \right] &= \sum_{b \in \mathcal{B}^N} p(\theta | X, b) p(b | A, X) \\
	&= \sum_{b \in \mathcal{B}^N} p(\theta, b | A, X) \\
	&= p(\theta | A, X) \\
\end{align*}
%
Which is indeed the distribution we are targeting from equation \ref{eqn:theta-target}. The reason we split the Markov chain into two stages is because the summation over all latent states $b \in \mathcal{B}^N$ required to compute the likelihood $p(A| X, \theta) = \sum_{b \in \Bcal^N}$ is computationally intensive

\subsection{Inferring block memberships}

\citet{Peixoto-MCMC} proposes a Monte Carlo method which we will base our approach on. It relies on writing the posterior in the following form.
%
\begin{equation}
	p(b | A, X) \propto p(A | b, X) \cdot p(b | X) = \pi_b(b)
\end{equation}
%
Now $\pi_b(\cdot)$ is the un-normalised density we wish to sample from. We can break $\pi_b$ down as follows:
%
\begin{align*}
	\pi_b(b) &= p(b|X) \sum_{\psi} \nolimits p(A , \psi | b, X) \\
	&= p(b|X) p(A, \psi^* | b, X) \\
	&= p(A | b, \psi^*) \cdot p(\psi^* | b) \cdot p(b | X)
\end{align*}
%
Since we are using a microcanonical formulation, there is only one value of $\psi$ which we denote $\psi^*$ that is compatible with the given $A$ and $b$. Namely, $k^*_i = \sum_j A_{ij} $ and $e^*_{rs} = \sum_{i, j} A_ij \one \{b_i=r\} \one\{b_j=s\}$. Therefore, the summation over all $\psi$ reduces to just one term that has the value $\psi^*$ compatible with $b$ and $\theta$. We also define the description length or entropy of the configuration as.
%
\begin{equation}
	S(b) = - \log \pi_b(b) = - \left[ \log p(A | b, \psi^*) + \log p(\psi^*, b | X) \right]
\end{equation}
%
Now this entropy can be thought of as the description length of the graph $A$ because it is the sum of the information required to represent the graph given the parameters and the amount of information required to store the parameters given $X$.

\subsection{Inferring feature to block generator}
